Instructions for developing new pages for the MarineOmics webpage

### 2023-12-14 DRAFT_DNA_methylation.Rmd
Update DRAFT_DNA_methylation.Rmd
added more context for bowtie

The purpose of this tutorial is to provide an example of aligning bisulfite-treated and enzymatically converted DNA sequence data and discuss needs for potential downstream applications. The principle behind creating bisulfite-treated DNA libraries is that when cytosines are not methylated, then they will be converted to uracil, which ultimately result in thymine nucleotides in the sequence data. @https://doi.org/10.1111/1755-0998.13542 offers a comparison of three methods for quantifying DNA methylation at single base-pair resolution using whole genome bisulfite sequencing (WGBS), reduced representation bisulfite sequencing (RRBS), and methyl-CpG binding domain bisulfite sequencing (MBDBS). When aligning bisulfite-converted data to a reference genome the percent methylation at a given CpG loci is determined by examining the ratio of thymines to cytosines. For example if there is a given cytosine adjacent to a guanine (ie CpG locus) with 30% of the reads containing thymine, one would consider this cytosine loci (CpG) to be 70% methylated. Since bisulfite treatment can damage DNA, newer library preparation and sequencing methods (ex. [NEBNext® Enzymatic Methyl-seq Kit; EM-seq](https://www.neb.com/en-us/products/e7120-nebnext-enzymatic-methyl-seq-kit#Protocols,%20Manuals%20&%20Usage) are starting to gain popularity. As EM-Seq produces loci-level data, alignment of EM-Seq data is similar to WGBS. Also, technologies like Nanopore and PacBio sequencing can directly detect DNA methylation without bisulfite conversion. As DNA passes through a nanopore or is sequenced by single-molecule real-time (SMRT) technology, changes in electrical current or fluorescence patterns can indicate the presence of methylated bases.

@@

## Considerations

The software used as part of this tutorial is **Bismark Bisulfite Mapper** (@10.1093/bioinformatics/btr167), which is one of the more widely-used alignment softwares for methylation data. Another commonly-used options include bwa-meth (@https://doi.org/10.48550/arXiv.1401.1129) and BSMAP (@https://doi.org/10.1186/1471-2105-10-232) (see @https://doi.org/10.1093/bib/bbab021 and @https://doi.org/10.1016/j.csbj.2022.08.051 for comparisons of these software for plant and mammal species, respectively). As with any software it is best to be famililar with the [User Manual](https://rawgit.com/FelixKrueger/Bismark/master/Docs/Bismark_User_Guide.html). The content below will provided with the assumption that the reader has read the manual and is meant to serve as guidance based on experience working with marine invertebrates.

The first step in the process in preparing the genome. Example code:

@@

    -   **`--parallel 28`**: This option tells Bismark to use 28 threads in parallel to speed up the process.

@@

>>>>

The purpose of this tutorial is to provide an example of aligning bisulfite-treated and enzymatically converted DNA sequence data to an existing genome, and discuss needs for potential downstream applications. The principle behind creating bisulfite-treated DNA libraries is that when cytosines are not methylated, then they will be converted to uracil, which ultimately result in thymine nucleotides in the sequence data. @https://doi.org/10.1111/1755-0998.13542 offers a comparison of three methods for quantifying DNA methylation at single base-pair resolution using whole genome bisulfite sequencing (WGBS), reduced representation bisulfite sequencing (RRBS), and methyl-CpG binding domain bisulfite sequencing (MBDBS). When aligning bisulfite-converted data to a reference genome the percent methylation at a given CpG loci is determined by examining the ratio of thymines to cytosines. For example if there is a given cytosine adjacent to a guanine (ie CpG locus) with 30% of the reads containing thymine, one would consider this cytosine loci (CpG) to be 70% methylated. Since bisulfite treatment can damage DNA, newer library preparation and sequencing methods (ex. [NEBNext® Enzymatic Methyl-seq Kit; EM-seq](https://www.neb.com/en-us/products/e7120-nebnext-enzymatic-methyl-seq-kit#Protocols,%20Manuals%20&%20Usage) are starting to gain popularity. As EM-Seq produces loci-level data, alignment of EM-Seq data is similar to WGBS. Also, technologies like Nanopore and PacBio sequencing can directly detect DNA methylation without bisulfite conversion. As DNA passes through a nanopore or is sequenced by single-molecule real-time (SMRT) technology, changes in electrical current or fluorescence patterns can indicate the presence of methylated bases.

@@

## Software Considerations

The software used as part of this tutorial is **Bismark Bisulfite Mapper** (@10.1093/bioinformatics/btr167), which is one of the more widely-used softwares to align methylation data to a reference genome. Another commonly-used options include bwa-meth (@https://doi.org/10.48550/arXiv.1401.1129) and BSMAP (@https://doi.org/10.1186/1471-2105-10-232) (see @https://doi.org/10.1093/bib/bbab021 and @https://doi.org/10.1016/j.csbj.2022.08.051 for comparisons of these software for plant and mammal species, respectively). As with any software it is best to be famililar with the [User Manual](https://felixkrueger.github.io/Bismark/).

Within Bismark, the tutorial uses Bowtie 2 as the alignment software.

The content below will provided with the assumption that the reader has read the manual and is meant to serve as guidance based on experience working with marine invertebrates.

## Genome Preparation

The first step in the process in preparing the reference genome by performing an in silico bisulfite conversion to allow Bowtie alignments. Example code:

@@

    -   **`--parallel 28`**: This option tells Bismark to use 28 threads in parallel to speed up the process. You can adjust this based on your computing power.

@@

## Aligning reads
